---
title: "Campaign Donations"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
```

The 2020 election was the most expensive in American history: Campaign spending on the presidential election and congressional races totaled a staggering 14 billion dollars, shattering the previous record of 7 billion dollars. The increasingly important role that money plays in American politics suggests some natural questions: Where does this money come from? What's driving the explosion in fundraising? How are fundraising totals associated with election results?

Our report attempts to (partially) answer these questions. We use data from the Federal Election Commission, henceforth abbreviated FEC, to look at trends in individual contributions to presidential campaigns over the past 20 years. Fortunately, the FEC website contains a treasure trove of detailed information, outlining how much money each candidate raised from each state, as well as how much they raised from five different donation amounts: donations under \$200, between \$200 and \$500, between \$500 and \$1000, between \$1000 and \$2000, and over \$2000.

```{r, echo = FALSE}

#Vector used to transform nominal values into 2020 dollars.
v <- c(1.5, 1.36, 1.18, 1.12, 1.07, 1)

campaignDonations <- read_csv(here("data", "CampaignFinanceDataOriginal.csv")) %>%
  mutate(total= under200 + between200and500 + between500and1000 + between1000and2000+ morethan2000) %>%
  mutate(real_total = total * v[(year - 1996)/4]) %>%
  mutate(party = case_when(
    candidate == "Joe Biden" ~ "D",
    candidate == "Donald Trump" ~ "R",
    candidate == "Hillary Clinton" ~ "D",
    candidate == "Barack Obama" ~ "D",
    candidate == "Mitt Romney" ~ "R",
    candidate == "John McCain" ~ "R",
    candidate == "George Bush" ~ "R",
    candidate == "John Kerry" ~ "D",
    candidate == "Al Gore" ~ "D"
  ))

write.csv(campaignDonations, "data/campaignDonations.csv")

totalDonations <- campaignDonations %>%
  group_by(year) %>%
  summarize(
    under200 = sum(under200),
    between200and500 = sum(between200and500),
    between500and1000 = sum(between500and1000),
    between1000and2000 = sum(between1000and2000),
    morethan2000 = sum(morethan2000),
    total = sum(total),
    real_total = sum(real_total)
  ) %>%
  pivot_longer(
    cols = c(under200, between200and500, between500and1000, between1000and2000, morethan2000),
    names_to = "type",
    values_to = "amount")

```

We first investigate changing patterns in campaign donations. Specifically, do individual donors give more money to Democratic presidential candidates or Republican presidential candidates? The following chart shows inflation-adjusted data from the past 20 years.

```{r, echo = FALSE}

ggplot(data = campaignDonations, aes(x = year, y = real_total/1000000)) +
  geom_line(aes(color = party)) +
  scale_color_manual(name = "Party", breaks = c("D", "R"), labels = c("Democrat", "Republican"), 
                     values = c("blue", "darkred")) +
  scale_x_continuous(breaks = seq(2000, 2020, by = 4)) +
  scale_y_continuous(breaks = seq(0, 1200, by = 100)) +
  xlab("Year") +
  ylab("Total Individual Contributions (Millions of Dollars)") +
  ggtitle("Total Individual Contributions by Party")

```

The diagram reveals a clear trend: While George Bush outraised his Democratic rivals in 2000 and 2004, the Democratic presidential candidate has enjoyed a large and growing fundraising advantage in every election since. While both parties have increased their fundraising levels from individual contributions, the trend has been driven primarily by increased donations to Democrats. One caveat applies, however: This is limited to individual contributions and does not include the fundraising of large Political Action Committees (PACs) that support or oppose particular candidates. PACs face far more lenient regulations -- some don't have to disclose the source of donations, and so-called super PACs can take unlimited money from corporations and labor unions -- so they have a large impact in elections. But perhaps the Democrats' large advantage in individual contributions is one reason why they want to limit the influence of PACs and why Republicans are more supportive of them.

In addition to looking at donations by party, we want to look at the total raised from each of the FEC's five donation amounts in each election. Has the increase in fundraising been driven by more large donations? More small donations? Donations in between? The following chart gives a clear answer.

```{r, echo = FALSE}

ggplot(data = totalDonations, aes(x = year, y = amount/1000000)) +
  geom_line(aes(color = type)) +
  scale_color_discrete(name = "Donation Type", 
                       breaks = c("under200", "between200and500", "between500and1000", 
                                  "between1000and2000", "morethan2000"), 
                       labels = c("Under $200", "Between $200 and $500", 
                                  "Between $500 and $1000", "Between $1000 and $2000",
                                  "More than $2000")) +
  scale_x_continuous(breaks = seq(2000, 2020, by = 4)) +
  scale_y_continuous(breaks = seq(0, 1000, by = 100)) +
  xlab("Year") +
  ylab("Total Contributed (Millions of Dollars)") +
  ggtitle("Individual Contributions by Donation Type")

```

While donations between \$200 and \$2000 have increased slightly since 2000, large donations above \$2000 have seen a much sharper increase. But by far the largest increase -- one that dwarfs all the others -- has been the rise of small donations under $200. This suggests that more people have been donating money (if the same number were donating more money, we would probably see a sharper rise in larger donation amounts).

Again, though, a few limitations complicate matters. Someone could theoretically donate multiple $200 checks, which would all fall in the "under \$200" category, when they really amount to larger donation. To correct for this, we would have to look at individual-level data. The FEC actually has datasets that show every single donation to each candidate, but these datasets are so large and so unwieldy that we didn't have the heart to use them. Additionally, we think that if someone wants to donate, say, \$1000, they would do so directly instead of writing five \$200 checks. Some people might have made multiple donations equal to one larger donation in a different bracket, but we think they represent a distinct minority of donors.

Inflation presents another problem. A \$900 donation in 2000 is the equivalent of about a \$1300 donation today. But the FEC classifies it in the "between \$500 and \$1000" category since the nominal value of the donation is \$900. As a result, simply adjusting the the total donations between \$500 and \$1000 in 2000 for inflation doesn't allow for an accurate comparison to that same number for the 2020 election. Some donations from 2000 that belong in a higher category (in 2020 dollars) would be counted in the lower donation bracket. We could adjust the FEC's five donation categories for inflation as well, so "donations under \$200" in 2000 would become "donations under \$300" in 2020 dollars (as the price level has risen about 50% in that time). But we still wouldn't get the desired comparison: donations in the 2000 election under \$200 in 2020 dollars versus donations in the 2020 election under \$200 in 2020 dollars. Once again, we could get around this with individual-level data, as we could adjust each individual donation for inflation and then compute the total for each donation bracket. Had we but world enough and time...

Given these issues, we've decided to leave the values in the chart in nominal terms (not adjusting for inflation). This is unfortunate, but there is one redeeming quality: Inflation has been quite low over the past 20 years, especially since the Great Recession. A dollar in 2008 is worth only \$1.18 today, so inflation alone cannot explain the dramatic increases in campaign donations.

What can? We speculate that the internet has made fundraising much easier and allowed candidates to reach a much larger pool of donors than they ever could before. Anyone with an internet connection can donate to a candidate in a manner of minutes: They need only go to the candidate's website, click one of the many bright buttons asking for financial support, and enter their payment details. At the same time, one could argue that the internet has heightened political polarization, causing more voters to donate to campaigns. But since our goal is statistical analysis and not amateur political punditry, we defer this idea to future researchers.

Having examined donation trends by party and donation level, we next combine both frameworks. In our Shiny app, the user can compare how much money selected candidates raised from specified donation categories. The same inflation issues apply, so we've left everything in nominal terms, but some interesting comparisons are still possible. For example, in 2012 Mitt Romney raised more money than Barack Obama from donations over \$2000. The 2016 election, however, saw a sharp reversal: Hillary Clinton raised more than Donald Trump from donations over \$2000. We invite you compare other candidates as well using the link below:

https://slawton21.shinyapps.io/Blog-Seamus-Matthew/



